---
title: "Price vs. Income: How (Un)affordable Has Housing Become?"
author: "Tobias van den Heuvel (2860237) 
Sem Veenhof (2856000)
Marnix Lakenman (2792960)
Diaz Terpstra (2782131)
Dennis Petrov (2765841)
Jan Wieger Overdijk (2782005)
Bilal el Ausrouti (2825152)"
date: "`r Sys.Date(23/06/2025)`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
# install.packages("tidyverse")
# install.packages("cbsodataR")
# install.packages("sf")
# install.packages("ggplot2")

```

```{r packages}
library(tidyverse)
library(cbsodataR)
library(sf)
library(ggplot2)
```

# Title Page

Tobias van den Heuvel (2860237) 
Sem Veenhof (2856000)
Marnix Lakenman (2792960)
Diaz Terpstra (2782131)
Dennis Petrov (2765841)
Jan Wieger Overdijk (2782005)
Bilal el Ausrouti (2825152)


Tutorial group 1
J.F.Fitzgerald


# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

The rising housing prices in the Netherlands have become a significant social problem, as the costs make it incredibly difficult for young adults to leave their parent’s homes.Between 2015-16 and 2021-22, the proportion of young people (18-34 years) able to move out within two years fell from 51% to 45% (Statistics Netherlands, 2024a). In 2024 the average rents rose by approximately 5 - 5.7%, which is the steepest jump since 1993, with social housing rents climbing by 5.6 - 5.7% (Statistics Netherlands, 2024b).This sharp rise squeezes low and middle income households, reducing disposable income for other essentials. As housing costs rise faster than incomes, younger adults and lower income renters/buyers face worsening housing uncertainty. Overcrowding becomes more common, and many who are not able to afford purchasing a home are pushed into long waiting lists for social housing.

# Part 2 - Data Sourcing

## 2.1 Load in the data

Data set verkoopprijs:
https://opendata.cbs.nl/#/CBS/nl/dataset/83625NED/table?ts=1748853166332 
We derived the data of the income per municipality by doing the following on the website on CBS. First, open the link to the correct page, then select at the filter “Onderwerp” all of the subjects under the tab “Inkomen en vermogen” (you should see a blue checkmark in a white little box next to “Inkomen en vermogen” and not a blue box). Next, select for “Perioden” all periods. And for “Regio’s” select all of the municipalities (i.e. “gemeenten per provincie”). Then click on “Download” at the right side of the page, and download as a “CSV met statistische symbolen”, this way you have downloaded the excel file with the correct data.

Data set inkomen:
https://opendata.cbs.nl/#/CBS/nl/dataset/70072ned/table?ts=1748937008485 
We derived the data of the housing prices per municipality by doing the following on the website on CBS. First open the second link to the correct page, then select at the filter “Onderwerp” all of the subjects. Next, select for “Perioden” all periods (1995-2023). Then select for “Regio’s” all of the “gemeenten per provincie”, you should see a blue checkmark in a white box at, “gemeenten per provincie” and at every province. So make sure you really have selected all municipalities. Then click on “Download” at the right side of the page, and download as a “CSV met statistische symbolen”, this way you have downloaded the excel file with the correct data.

Data set bevolkingsdichtheid:
https://opendata.cbs.nl/statline/#/CBS/nl/dataset/70072ned/table?ts=1750083971495 
We derived the data of population density by doing the following on the website on CBS. First open the third link to the correct page, then select at the filter 
“Onderwerp” the box called “bevolkingsdichtheid”. To get to this box you have to deselect every subject (so no subjects are selected). then, click on the arrow at “Bevolking”, then on the arrow of “Bevolkingssamenstelling op 1 januari” then choose “bevolkingsdichtheid”. Make sure only bevolkingsdichtheid is selected. Next, select for “Perioden” all periods (1995-2023). Then select for “Regio’s” all of the “gemeenten per provincie”, you should see a blue checkmark in a white box at, “gemeenten per provincie” and at every province. So make sure you really have selected all municipalities. Then click on “Download” at the right side of the page, and download as a “CSV met statistische symbolen”, this way you have downloaded the excel file with the correct data.


```{r loading_data}
data_set_huizenprijzen <- read.csv2("data/Huisprijzen.csv")
data_set_inkomen <- read.csv2("data/Regionale_kerncijfers_Nederland_04062025_092105.csv")

data_set_bevolkingsdichtheid <- read.csv("data/Regionale_kerncijfers_Nederland_16062025_162909.csv")
data_set_bevolkingsdichtheid <- data_set_bevolkingsdichtheid %>%
  separate(
    col = Perioden.Regio.s.Bevolking.Bevolkingssamenstelling.op.1.januari.Bevolkingsdichtheid..aantal.inwoners.per.km..,
    into = c("Perioden", "Regio.s", "Bevolkingsdichtheid"),
    sep = ";"
  ) %>%
  mutate(
    Perioden = as.integer(Perioden),
    Regio.s = trimws(Regio.s),
    Bevolkingsdichtheid = as.numeric(Bevolkingsdichtheid)
  ) %>%
  filter(!is.na(Bevolkingsdichtheid))
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(subset1)
head(subset2)
head(data_set_bevolkingsdichtheid)
```
The first dataset is from CBS, contains data about the mean income of a household per municipality and is called “Regionale kerncijfers Nederland”. The data ranges from 1995 to 2023, however not every year and region has all of the data, but most of the regions and years have.
 The second dataset from CBS, contains data about the selling prices of houses per municipality and is called “Bestaande koopwoningen; gemiddelde verkoopprijzen, regio”, the data ranges from 1995 to 2023, however not every year and region have all data, but most of the regions and years have.
 The Third dataset contains data about the population density in every municipality in the Netherlands. Also Here the data is derived from CBS and is called “Regionale kerncijfers Nederland”.

``` r
inline_code = TRUE
```

## 2.3 Describe the type of variables included

We use two datasets from Statistics Netherlands (CBS). The first dataset provides information on the average selling price of houses per municipality, while the second covers average disposable income per household.
Both datasets include the following types of variables:
Categorical variables:
 Municipality (e.g., Amsterdam, Utrecht)
 Year (e.g., 2019, 2020)


Numerical variables:
 Average selling price per dwelling (in euros)
 Average disposable income per household (in thousands of euros)
 Number of dwellings / households


These variables allow us to analyze trends over time and compare municipalities in terms of property values and income levels.


# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
subset1 <- data_set_huizenprijzen[, 1:3]

# Pak kolommen 1, 2 en 4 uit inkomen, bijvoorbeeld: Regio.s, Perioden, Inkomen
subset2 <- data_set_inkomen[, c(1, 2, 4)]

# Hernoem kolom 3 in subset2 naar Inkomen 
colnames(subset2)[3] <- "Inkomen"

# Merge op Regio.s en Perioden
merged_data <- merge(subset1, subset2, by = c("Regio.s", "Perioden"))

# Verwijder rijen met missende waarden
clean_data <- na.omit(merged_data)

#Blank cells will become NA's
clean_data$Inkomen=as.numeric(clean_data$Inkomen)

#gemiddelde verkoopprijs delen door 1000
clean_data$Gemiddelde.verkoopprijs..euro. = clean_data$Gemiddelde.verkoopprijs..euro. / 1000

#verwijderen van rijen met missende data 
clean_data <- na.omit(clean_data)

```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
clean_data$`Prijs Inkomensratio` <- clean_data[[3]] / clean_data[[4]]
names(clean_data)[names(clean_data) == "Prijs Inkomensratio"] <- "Prijs.Inkomensratio"
```

Variable 2

```{r gen_var2}
clean_data <- clean_data %>%
  arrange(Regio.s, Perioden) %>%
  group_by(Regio.s) %>%
  mutate(
    Groei_Verkoopprijs = (Gemiddelde.verkoopprijs..euro. - lag(Gemiddelde.verkoopprijs..euro.)) / lag(Gemiddelde.verkoopprijs..euro.),
    Groei_Inkomen = (Inkomen - lag(Inkomen)) / lag(Inkomen),
    Groei_PIR = (Prijs.Inkomensratio - lag(Prijs.Inkomensratio)) / lag(Prijs.Inkomensratio)
  ) %>%
  ungroup()
```

## 3.3 Visualize temporal variation

```{r}
# grafiek groei PIR
gemiddelde_groei_per_jaar <- clean_data %>%
  group_by(Perioden) %>%
  summarise(gemiddelde_groei = mean(Groei_PIR, na.rm = TRUE))

ggplot(gemiddelde_groei_per_jaar, aes(x = Perioden, y = gemiddelde_groei)) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "grey", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "black", size = 0.8) +
  labs(title = "Average growth of the PIR per Year\n",
       x = "Year\n",
       y = "Average growth\n") +
  scale_x_continuous ("Year", breaks = 2011:2023) +
  scale_y_continuous ("Average growth", lim= c(-0.12,0.12)) +
  theme_minimal()


```
The temporal analysis of the average growth of the Price-to-Income Ratio (PIR) per year illustrates how the relationship between housing prices and average household income has evolved over time. This analysis is important because the PIR directly reflects housing affordability. An increasing PIR indicates that housing prices are rising faster than incomes, making it more difficult for households to afford a home. The graph shows significant increases in the PIR during the years 2015, 2020, and 2021, suggesting a decline in affordability during those periods. In contrast, the sharp drop in 2022 and 2023 may point to a price correction or income growth. By visualizing these annual fluctuations, the analysis highlights when imbalances between housing prices and income emerged, which is highly relevant to showcase the social problem.

## 3.4 Visualize spatial variation

```{r visualise_map}
#nieuwe tabel aangemaakt voor shapefiles
data_2019 <- clean_data %>%
  filter(Perioden == 2019) %>%
  select(Regio.s, PIR_2019 = 5)

data_2023 <- clean_data %>%
  filter(Perioden == 2023) %>%
  select(Regio.s, PIR_2023 = 5)

pir_data <- merge(data_2019, data_2023, by = "Regio.s")

#laad shapefile
gemeente_map <- cbs_get_sf("gemeente", 2023, verbose = TRUE)
head(gemeente_map)

#een nieuwe dataset maken waarbij de juiste data word samengevoegd voor de shapefile om te werken
colnames(pir_data)[which(colnames(pir_data) == "Regio.s")] <- "statnaam"
kaart_met_data <- left_join(gemeente_map, pir_data, by = "statnaam")

ggplot(kaart_met_data) +
  geom_sf(aes(fill = PIR_2019), color = "white") +
  scale_fill_viridis_c(option = "C", na.value = "grey90", breaks = seq(5, 10, by = 1), lim = c(4, 11)) +
  labs(title = "Price/Income Ratio per Municipality (PIR) (2019)\n", fill = "PIR") +
  theme_minimal()

ggplot(kaart_met_data) +
  geom_sf(aes(fill = PIR_2023), color = "white") +
  scale_fill_viridis_c(option = "C", na.value = "grey90", breaks = seq(5, 10, by = 1), lim = c(4, 11)) +
  labs(title = "Price/Income Ratio per Municipality (PIR) (2023)\n", fill = "PIR") +
  theme_minimal()

```

These shapefiles, depicting the price by income ratio (PIR) in 2019 and 2023 in different municipalities, are highly relevant to our social problem to provide a good view of the income differences and house prices in all municipalities in the Netherlands. It clearly shows that in some municipalities it could be easier to buy a home than in other municipalities, that is because the mean salary and the price of homes of all municipalities in the Netherlands are different.
The growth of the PIR across 2019 to 2023 also shows how the ratio changed during these years, which had a big impact on housing affordability and income. This clearly helps with our social problem, because with these shapefiles we are able to clearly depict the social problem on a map and therefore the right measures can be taken on a more specific level, namely on municipality level. 


## 3.5 Visualize sub-population variation

What is the PIR per poppulation desinty quintile?

```{r visualise_subpopulations}
df_quintielen <- data_set_bevolkingsdichtheid %>%
  filter(!is.na(Bevolkingsdichtheid)) %>%               
  group_by(Perioden) %>%                                
  mutate(Bevolkingsdichtheid_quintiel = ntile(Bevolkingsdichtheid, 5)) %>%
  ungroup()

df_quintielen <- df_quintielen %>%
  mutate(density_group = case_when(
    Bevolkingsdichtheid_quintiel == 1 ~ "Very low density",
    Bevolkingsdichtheid_quintiel == 2 ~ "Low density",
    Bevolkingsdichtheid_quintiel == 3 ~ "Medium density",
    Bevolkingsdichtheid_quintiel == 4 ~ "High density",
    Bevolkingsdichtheid_quintiel == 5 ~ "Very high density"
  ))

# Zet juiste factorvolgorde 
df_quintielen$density_group <- factor(df_quintielen$density_group,
                                   levels = c("Very low density", "Low density",
                                              "Medium density", "High density",
                                              "Very high density"))

# join de datasets
df_complete <- clean_data %>%
  left_join(df_quintielen %>% select(Regio.s, Perioden, density_group),
            by = c("Regio.s", "Perioden"))

#Boxplot voor 2019 en 2023
ggplot(df_complete %>% filter(Perioden %in% c(2019, 2023)),
       aes(x = density_group, y = Prijs.Inkomensratio)) +
  geom_boxplot(aes(fill = as.factor(Perioden))) +
  labs(
    title = "Price-to-Income Ratio by Population Density Group (2019 vs 2023)\n",
    x = "Quintiles of Population Density\n",
    y = "Price-to-Income Ratio\n",
    fill = "Year"
  ) +
  theme_minimal()
```

This boxplot, showing the price-to-income ratio (PIR) across different population density groups in both 2019 and 2023, is highly relevant to our social problem as it illustrates how housing affordability has developed over time in areas with varying levels of urbanization. It gives a clear overview of how much more difficult it has become to afford a house, especially in municipalities with higher population densities.
The comparison between 2019 and 2023 reveals that the PIR has increased in all quintiles of population density, which means that house prices have grown faster than incomes in every group. However, the increase is most noticeable in medium to very high-density areas, where the affordability gap is becoming more severe. This reflects the pressure on urban housing markets and shows that people living in or moving to cities are being disproportionately affected.
This data is important for our research, because it allows us to link population density directly to the affordability issue we are analyzing. It also supports our argument that the housing problem is not the same in every part of the country and that some areas need more targeted policy measures. By using this breakdown by density, we can better understand where affordability is declining the fastest and where interventions are most urgently needed.
Overall, this figure adds depth to our analysis and strengthens the link between rising house prices, stagnant incomes, and increasing inequality in housing accessibility.


## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
# grafiek groei PIR
gemiddelde_groei_per_jaar <- clean_data %>%
  group_by(Perioden) %>%
  summarise(gemiddelde_groei = mean(Groei_PIR, na.rm = TRUE))

ggplot(gemiddelde_groei_per_jaar, aes(x = Perioden, y = gemiddelde_groei)) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "grey", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "black", size = 0.8) +
  geom_vline(xintercept = 2020, linetype = "dotted", color = "red", size = 1) +
  geom_vline(xintercept = 2022, linetype = "dotted", color = "red", size = 1) +
  labs(title = "Average growth of the PIR per Year\n",
       x = "Year\n",
       y = "Average growth\n") +
  scale_x_continuous ("Year", breaks = 2011:2023) +
  scale_y_continuous ("Average growth", lim= c(-0.12,0.12)) +
  theme_minimal()

#grafiek groei gemiddelde inkomen
gemiddelde_groei_inkomen_per_jaar <- clean_data %>%
  group_by(Perioden) %>%
  summarise(gemiddelde_groei = mean(Groei_Inkomen, na.rm = TRUE))

ggplot(gemiddelde_groei_inkomen_per_jaar, aes(x = Perioden, y = gemiddelde_groei)) +
  geom_line(color = "blue", size = 1) +
    geom_smooth(method = "lm", se = FALSE, color = "grey", linetype = "dashed") +
  geom_hline(yintercept = 0, color = "black", size = 0.8) +
  labs(title = "Average growth of the income per Year\n",
       x = "Year\n",
       y = "Average growth\n") +
  scale_x_continuous ("Year", breaks = 2011:2023) +
  scale_y_continuous ("Average growth") +
  theme_minimal()
```

The coronavirus pandemic significantly intensified housing affordability issues, as seen by the sharp rise in the PIR from 2020 to 2022. This was driven by surging house prices, while income growth lagged, especially early in the pandemic. In times of economic concerns people fear their savings being eaten up by inflation and see purchasing properties as a safe investment. Lockdown and remote work further led more people into buying homes -. the prices in large urban municipalities continued to increase exponentially, however prices in smaller or rural municipalities also saw a drastic increase as people wanted to invest into more livable space (Klava, 2024).

The average growth of income per year plummeted in 2020, as companies halted all recruitment and were hesitant to promote internally, possibly affected by the economic uncertainties. In 2023 a correction began and the income growth started recovering to pre-2020 numbers, but PIR remained elevated, particularly in urban and densely populated areas.


# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: https://github.com/Sem2909/Group-Project-ProgrammingForEcon-group-5

## 5.2 Reference list

Centraal Bureau voor de Statistiek. (2025, May 22). Bestaande koopwoningen; verkoopprijzen prijsindex 2020=100. https://opendata.cbs.nl/#/CBS/nl/dataset/85773NED/table?ts=1729589594896

Centraal Bureau voor de Statistiek. (2025, February 17). Bestaande koopwoningen; gemiddelde verkoopprijzen, regio. https://opendata.cbs.nl/#/CBS/nl/dataset/83625NED/table?ts=1748853166332

CBS Statline. (2, June, 2025.-a).
https://opendata.cbs.nl/statline/#/CBS/nl/dataset/70072ned/table?ts=1750083971495

CBS Statline. (2, June, 2025.-b)
https://opendata.cbs.nl/#/CBS/nl/dataset/70072ned/table?ts=1748937008485 

CBS Statline. (17, Febuary, 2025.-c).
https://opendata.cbs.nl/#/CBS/nl/dataset/83625NED/table?ts=1748853166332

CBS Statistics Netherlands. (2024, June 19). Fewer young people are able to leave parental home.https://www.cbs.nl/en-gb/news/2024/25/fewer-young-people-are-able-to-leave-parental-home

CBS Statistics Netherlands. (2024b, September 4). Rent increases the highest in over 30 years.https://www.cbs.nl/en-gb/news/2024/36/rent-increases-the-highest-in-over-30-years

Klava, A. (2024, June). Housing Demand in a Work-From-Home Economy (Master’s Thesis, Vrije University Amsterdam)
https://spatialeconomics.nl/wp-content/uploads/2024/10/STR_2024_KlavaA_Housing_Demand_in_a_Work_from_Home_Economy.pdf


